Binning Strategies and Related Loss for Large Data

### 2015-04-10 Paper.Rnw
"L" used to both denote loss and index 2d bins.  change to uppercase script L to match \ell notation.  Also get rid of "+1" to bin counts in log freq loss since we can simply change notation to sum over only, non-empty bins.  This in turn makes the log frequency transformation for binning more simple

\documentclass[letterpaper,10pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amsthm,amssymb}
\usepackage{fullpage}
\usepackage{graphicx,float,wrapfig,subfig,tabularx,ulem}
\graphicspath{{figure/}}
%\usepackage{csquotes}
\usepackage{color}
\usepackage{natbib}
\usepackage{hyperref}

@@

The standard and random binning algorithms are easily extendable to higher dimension. For the purposes of creating binned scatterplots we will specify extension to rectangular binning in two dimensions. In this case we wish to assign data pairs $(x_i, y_i)$ to bin centers of the form $(x_j^\ast,y_k^\ast)$, with $j \in \{1,\dots,J\}$ and $k \in \{1,\dots,K\}$, where $J$ and $K$ are the number of bins in the X and Y dimensions, respectively. The two ($j$,$k$) pairs that index the bin centers can be linearized to a single index such that $\ell = j + J(k-1)$; thus making $j$ the fast running index and $k$ the slow running index.  With this linearized index for all bins we now have a set of bin centers of the form $(x_\ell^\ast,y_\ell^\ast)$, with $\ell \in \{1,\dots,L\}$, where $L=J\cdot K$.

@@

Loss of information occurs during the binning and rendering process. For the remainder of the paper we will assume that we are using colors in binned scatterplots to represent frequencies.  We distinguish two sources of loss in the construction of a binned scatterplot:
\begin{itemize}
\item {\it Spatial Loss}, $L^\text{S}$, occurs when points $(x_i, y_i)$ for observations $i \in \{1,\dots,n\}$ in the data set are reduced to a single points at the bin centers $(x_\ell^\ast, y_\ell^\ast)$ for bins $\ell \in \{1,\dots,L\}$. By displaying frequency information using tiles that are visually centered at the bin centers causes a loss of information about the exact location of the points.

@@

 Whenever the shading scheme for rendering the counts in each bin is discretized there is a loss of frequency information. We can model this as a second stage of binning; wherein the bin counts, $c_\ell$, for bins $\ell \in \{1, \dots, L \}$ are placed into frequency bins using any of the previously discussed univariate binning algorithms in Section~\ref{GenBinning}.  Figure~\ref{fig:FreqLossDemo} provides a visual example of a discrete color palette resulting from frequency binning.

@@

The goal of binning the frequencies and using a ordinally shaded color palette is to provide more distinctive perceivable breaks to aid the information extraction; however it does so at the cost of using shade to display binned frequencies, $b_c(c_\ell)$ instead of the true frequencies, $c_\ell$. The \textit{total frequency loss}, $L^F$, is defined as
%
\begin{eqnarray}\label{totfreqloss}
L^F = \sum_{\ell=1}^{L} L_\ell^F = \sum_{\ell=1}^{L} (c_\ell-b_c(c_\ell))^2
\end{eqnarray}  
%
where $L_\ell^F$ is the frequency loss for the $\ell^{th}$ bin. Note that this is effectively a sum of squared error from binning frequencies.

@@

Another alternative is to use a log transformation which produces a more symmetric distribution of frequency information, increasing perceptual resolution. This is consistent with the Weber-Fechner law which suggests that increased stimulus intensity is perceptually mapped on the log scale \citep{sp}. Using a logarithmic mapping of frequency to the color or size aesthetic provides a more natural perceptual experience and simultaneously increases the perceptual resolution of the graph. The \textit{log frequency loss}, $L^{\log F}$, is defined as
%
\begin{eqnarray}\label{logfrequency}
L^{\log F} = \sum_{\ell=1}^{L} L_\ell^{\log F} = \sum_{\ell=1}^{L} (\log(c_\ell+1) - \log(b_c(c_\ell+1)))^2
\end{eqnarray}
%
Note that we have added one count to every cell before taking the logarithm to avoid problems with empty cells.

@@

\begin{figure}[hbtp]
  \begin{minipage}{.31\linewidth}
<<noFreqBin, echo=FALSE, out.width='\\linewidth',fig.width=4, fig.height=4>>=
binout1 <- RectBin2d(datlist[[2]]$x,datlist[[2]]$y,0,0,5,5,type="standard")
qplot(binxs, binys, geom="tile", fill=binfreq, data=binout1[[1]],
            main="No Freq Binning") +
  xlab("") + ylab("")+
  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43",
                      guide = guide_colourbar(label.position="bottom", label.hjust=0.5,  
                                           title.position = "top")) + 
  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
                     panel.grid= element_blank(),
                     plot.margin = unit(c(.1,.1,.1,.1),"cm") )
@
  \end{minipage}
\hfil\begin{minipage}{.685\linewidth}
<<FreqBinPlots, echo=FALSE, out.width='\\linewidth',fig.width=10, fig.height=10>>=
binoutFreqGroups <- freqBin(binout1, binType="standard", ncolor=5, logCount=FALSE)
binoutQuantFreqGroups <- freqBin(binout1, binType="quantile", ncolor=5, logCount=FALSE)
binoutLogFreqGroups <- freqBin(binout1, binType="standard", ncolor=5, logCount=TRUE)
binoutQuantLogFreqGroups <- freqBin(binout1, binType="quantile", ncolor=5, logCount=TRUE)

# Plot with Binned Frequencies
labels = paste(unique(binoutFreqGroups[[1]]$freqlabel), "\n(", unique(binoutFreqGroups[[1]]$freqgroup), ")", sep="")
p2 <- qplot(binxs, binys, geom="tile", fill=freqgroup, data=binoutFreqGroups[[1]],
            main="Standard Freq Binning") +
    xlab("") + ylab("")+
  scale_fill_gradient("Frequencies", low="#56B1F7", high="#132B43", 
                      guide = guide_legend(label.position="bottom", label.hjust=0.5,  
                                           title.position = "top"),
                      breaks=unique(binoutFreqGroups[[1]]$freqgroup),
                      labels=labels) + 
  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
                     legend.key.width = unit(1, "cm"),
                     panel.grid= element_blank(),
                     plot.margin = unit(c(0.5,.1,.1,.1),"cm"),
                     legend.text=element_text(size=12) )

# Plot with Quantile Binned Frequencies
labels = paste(unique(binoutQuantFreqGroups[[1]]$freqlabel), "\n(", unique(binoutQuantFreqGroups[[1]]$freqgroup), ")", sep="")
cols <- seq_gradient_pal(low="#56B1F7", high="#132B43", space="Lab")((0:5)/5)
p3 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup), data=binoutQuantFreqGroups[[1]],
            main="Quantile Freq Binning") +
    xlab("") + ylab("")+
  scale_fill_manual("Frequencies", values=cols, 
                      guide = guide_legend(label.position="bottom", label.hjust=0.5,  
                                           title.position = "top"),
                      labels=labels) + 
  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
                     legend.key.width = unit(1, "cm"),
                     panel.grid= element_blank(),
                     plot.margin = unit(c(0.5,.1,.1,.1),"cm"),
                     legend.text=element_text(size=12) ) 
# Plot with standard binned Log Frequencies
labels = paste(unique(binoutLogFreqGroups[[1]]$freqlabel), "\n(", unique(binoutLogFreqGroups[[1]]$freqgroup), ")", sep="")
cols <- seq_gradient_pal(low="#56B1F7", high="#132B43", space="Lab")((0:5)/5)
p4 <- qplot(binxs, binys, geom="tile", fill=freqgroup, data=binoutLogFreqGroups[[1]],
            main="Standard Log Freq Binning") +
    xlab("") + ylab("")+
  scale_fill_gradient("Log Frequencies", low="#56B1F7", high="#132B43", 
                      guide = guide_legend(label.position="bottom", label.hjust=0.5,  
                                           title.position = "top"),
                      breaks=unique(binoutLogFreqGroups[[1]]$freqgroup),
                      labels=labels) + 
  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
                     legend.key.width = unit(1, "cm"),
                     panel.grid= element_blank(),
                     plot.margin = unit(c(0.5,.1,.1,.1),"cm") ,
                     legend.text=element_text(size=12))

# Plot with quantile binned Log Frequencies
labels = paste(unique(binoutQuantLogFreqGroups[[1]]$freqlabel), "\n(", unique(binoutQuantLogFreqGroups[[1]]$freqgroup), ")", sep="")
p5 <- qplot(binxs, binys, geom="tile", fill=freqgroup,
            data=binoutQuantLogFreqGroups[[1]],
            main="Quantile Log Freq Binning") +
    xlab("") + ylab("")+
  scale_fill_gradient("Log Frequencies", low="#56B1F7", high="#132B43", 
                      guide = guide_legend(label.position="bottom", label.hjust=0.5,  
                                           title.position = "top"),
                      breaks=unique(binoutQuantLogFreqGroups[[1]]$freqgroup),
                      labels=labels) + 
  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
                     legend.key.width = unit(1, "cm"),
                     panel.grid= element_blank(),
                     plot.margin = unit(c(0.5,.1,.1,.1),"cm"),
                     legend.text=element_text(size=12))

>>>>

\documentclass[letterpaper,10pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amsthm,amssymb}
\usepackage{fullpage}
\usepackage{graphicx,float,wrapfig,subfig,tabularx,ulem}
\graphicspath{{figure/}}
%\usepackage{csquotes}
\usepackage{color}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{mathrsfs}

@@

The standard and random binning algorithms are easily extendable to higher dimension. For the purposes of creating binned scatterplots we will specify extension to rectangular binning in two dimensions. In this case we wish to assign data pairs $(x_i, y_i)$ to bin centers of the form $(x_j^\ast,y_k^\ast)$, with $j \in \{1,\dots,J\}$ and $k \in \{1,\dots,K\}$, where $J$ and $K$ are the number of bins in the X and Y dimensions, respectively. The two ($j$,$k$) pairs that index the bin centers can be linearized to a single index such that $\ell = j + J(k-1)$; thus making $j$ the fast running index and $k$ the slow running index.  With this linearized index for all bins we now have a set of bin centers of the form $(x_\ell^\ast,y_\ell^\ast)$, with $\ell \in \{1,\dots,\mathscr{L}\}$, where $\mathscr{L}=J\cdot K$.

@@

Loss of information occurs during the binning and rendering process. For the remainder of the paper we will assume that we are using colors in binned scatterplots to represent frequencies.  We distinguish two sources of loss in the construction of a binned scatterplot:
\begin{itemize}
\item {\it Spatial Loss}, $L^\text{S}$, occurs when points $(x_i, y_i)$ for observations $i \in \{1,\dots,n\}$ in the data set are reduced to a single points at the bin centers $(x_\ell^\ast, y_\ell^\ast)$ for bins $\ell \in \{1,\dots,\mathscr{L}\}$. By displaying frequency information using tiles that are visually centered at the bin centers causes a loss of information about the exact location of the points.

@@

 Whenever the shading scheme for rendering the counts in each bin is discretized there is a loss of frequency information. We can model this as a second stage of binning; wherein the bin counts, $c_\ell$, for bins $\ell \in \{1, \dots, \mathscr{L} \}$ are placed into frequency bins using any of the previously discussed univariate binning algorithms in Section~\ref{GenBinning}.  Figure~\ref{fig:FreqLossDemo} provides a visual example of a discrete color palette resulting from frequency binning.

@@

The goal of binning the frequencies and using a ordinally shaded color palette is to provide more distinctive perceivable breaks to aid the information extraction; however it does so at the cost of using shade to display binned frequencies, $b_c(c_\ell)$ instead of the true frequencies, $c_\ell$. The \textit{total frequency loss}, $L^F$, is defined as
%
\begin{eqnarray}\label{totfreqloss}
L^F = \sum_{\ell=1}^{\mathscr{L}} L_\ell^F = \sum_{\ell=1}^{\mathscr{L}} (c_\ell-b_c(c_\ell))^2
\end{eqnarray}  
%
where $L_\ell^F$ is the frequency loss for the $\ell^{th}$ bin. Note that this is effectively a sum of squared error from binning frequencies.

@@

Another alternative is to use a log transformation which produces a more symmetric distribution of frequency information, increasing perceptual resolution. This is consistent with the Weber-Fechner law which suggests that increased stimulus intensity is perceptually mapped on the log scale \citep{sp}. Using a logarithmic mapping of frequency to the color or size aesthetic provides a more natural perceptual experience and simultaneously increases the perceptual resolution of the graph. The \textit{log frequency loss}, $L^{\log F}$, is defined as
%
\begin{eqnarray}\label{logfrequency}
L^{\log F} = \sum_{\ell \in \mathscr{L}^\ast} L_\ell^{\log F} = \sum_{\ell \in \mathscr{L}^\ast} (\log(c_\ell) - \log(b_c(c_\ell)))^2
\end{eqnarray}
%
where $\mathscr{L}^\ast$ is the index set for all non-empty bins, which is done to avoid asymptotic problems from log transforming bin counts of zero.

@@

\begin{figure}[hbtp]
  \begin{minipage}{.31\linewidth}
<<noFreqBin, echo=FALSE, out.width='\\linewidth',fig.width=4, fig.height=4>>=
binout1 <- RectBin2d(datlist[[2]]$x,datlist[[2]]$y,0,0,5,5,type="standard")
qplot(binxs, binys, geom="tile", fill=binfreq, data=binout1[[1]],
            main="No Freq Binning") +
  xlab("") + ylab("")+
  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43",
                      guide = guide_colourbar(label.position="bottom", label.hjust=0.5,  
                                           title.position = "top")) + 
  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
                     panel.grid= element_blank(),
                     plot.margin = unit(c(.1,.1,.1,.1),"cm") )
@
  \end{minipage}
\hfil\begin{minipage}{.685\linewidth}
<<FreqBinPlots, echo=FALSE, out.width='\\linewidth',fig.width=10, fig.height=10>>=
freqgrps = 5
binoutFreqGroups <- freqBin(binout1, binType="standard", ncolor=freqgrps, logCount=FALSE)
binoutQuantFreqGroups <- freqBin(binout1, binType="quantile", ncolor=freqgrps, logCount=FALSE)
binoutLogFreqGroups <- freqBin(binout1, binType="standard", ncolor=freqgrps, logCount=TRUE)
binoutQuantLogFreqGroups <- freqBin(binout1, binType="quantile", ncolor=freqgrps, logCount=TRUE)
cols <- seq_gradient_pal(low="#56B1F7", high="#132B43", space="Lab")((0:freqgrps)/freqgrps)

# Plot with Binned Frequencies
labels = paste(unique(binoutFreqGroups[[1]]$freqlabel), "\n(", unique(binoutFreqGroups[[1]]$freqgroup), ")", sep="")
p2 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup), data=binoutFreqGroups[[1]],
            main="Standard Freq Binning") +
    xlab("") + ylab("")+
  scale_fill_manual("Frequencies", values=cols, 
                      guide = guide_legend(label.position="bottom", label.hjust=0.5,  
                                           title.position = "top"),
                      breaks=unique(binoutFreqGroups[[1]]$freqgroup),
                      labels=labels) + 
  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
                     legend.key.width = unit(1, "cm"),
                     panel.grid= element_blank(),
                     plot.margin = unit(c(0.5,.1,.1,.1),"cm"),
                     legend.text=element_text(size=12) )

# Plot with Quantile Binned Frequencies
labels = paste(unique(binoutQuantFreqGroups[[1]]$freqlabel), "\n(", unique(binoutQuantFreqGroups[[1]]$freqgroup), ")", sep="")
p3 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup), data=binoutQuantFreqGroups[[1]],
            main="Quantile Freq Binning") +
    xlab("") + ylab("")+
  scale_fill_manual("Frequencies", values=cols, 
                      guide = guide_legend(label.position="bottom", label.hjust=0.5,  
                                           title.position = "top"),
                      labels=labels) + 
  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
                     legend.key.width = unit(1, "cm"),
                     panel.grid= element_blank(),
                     plot.margin = unit(c(0.5,.1,.1,.1),"cm"),
                     legend.text=element_text(size=12) ) 
# Plot with standard binned Log Frequencies
labels = paste(unique(binoutLogFreqGroups[[1]]$freqlabel), "\n(", unique(binoutLogFreqGroups[[1]]$freqgroup), ")", sep="")
p4 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup), data=binoutLogFreqGroups[[1]],
            main="Standard Log Freq Binning") +
    xlab("") + ylab("")+
  scale_fill_manual("Log Frequencies", values=cols, 
                      guide = guide_legend(label.position="bottom", label.hjust=0.5,  
                                           title.position = "top"),
                      breaks=unique(binoutLogFreqGroups[[1]]$freqgroup),
                      labels=labels) + 
  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
                     legend.key.width = unit(1, "cm"),
                     panel.grid= element_blank(),
                     plot.margin = unit(c(0.5,.1,.1,.1),"cm") ,
                     legend.text=element_text(size=12))

# Plot with quantile binned Log Frequencies
labels = paste(unique(binoutQuantLogFreqGroups[[1]]$freqlabel), "\n(", unique(binoutQuantLogFreqGroups[[1]]$freqgroup), ")", sep="")
p5 <- qplot(binxs, binys, geom="tile", fill=factor(freqgroup),
            data=binoutQuantLogFreqGroups[[1]],
            main="Quantile Log Freq Binning") +
    xlab("") + ylab("")+
  scale_fill_manual("Log Frequencies", values=cols, 
                      guide = guide_legend(label.position="bottom", label.hjust=0.5,  
                                           title.position = "top"),
                      breaks=unique(binoutQuantLogFreqGroups[[1]]$freqgroup),
                      labels=labels) + 
  theme_bw() + theme(legend.position="bottom",aspect.ratio=1,
                     legend.key.width = unit(1, "cm"),
                     panel.grid= element_blank(),
                     plot.margin = unit(c(0.5,.1,.1,.1),"cm"),
                     legend.text=element_text(size=12))
